Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-21 Paper/Paper.tex
recent changes to the paper

Sensing a persons activity is being researched for several years now, raising even more interest in it by the advanced technology of mobile phones, which include multiple sensors and therefore enable activity recognition with such a device \cite{brezmes2009activity}. Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors embedded in a mobile device \cite{brezmes2009activity} or attached to the body \cite{Ravi2005} to identify the activities performed. Usually, single sensor modalities are used or duplicated to detect the activities. However, it is a great challenge to identify fine-grained activities just by using a single modality like the accelerometer. Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the 
object's distance, its size and the material it is made of. In this work, we show how an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living. For this we obtained a wrist-worn activity data logger\footnote{HedgeHog Activity Logger -- \url{http://www.ess.tu-darmstadt.de/hedgehog}} and integrate a capacitive proximity sensor unobtrusively into the wristband.

@@

\begin{figure}
	\centering
 		\includegraphics[trim=1cm 1cm 1cm 1cm,clip,width=\textwidth]{Images/pseudocode.pdf}
	\caption{Measurement procedure carried out by the HedgeHog sensor: using the microcontroller's Timer0 module in counting mode, the oscillating signal generated by the capacitive sensor circuit can be measured by counting the frequency pulses over a predefined gate time (approximately 9.5ms).}
	\label{fig:pseudocode}
\end{figure}

Figure \ref{fig:pseudocode} illustrates the operations required for a measurement cycle. The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms. During that counting phase, the HedgeHog is sent to sleep in order to reduce power consumption. In the following, the HedgeHog applies run-length encoding on the measured data to reduce overhead and periodically logs the data to the integrated SD card.

@@

The experiment setup aims to depict a typical scenario in daily life. Especially in the field of Ambient Assisted Living (AAL), it is desired to monitor activities like drinking, preparing lunch and sleeping. A fine-grained monitoring of such activities may help elderly or people suffering from mental diseases to maintain a healthy day/night rhythm and take action if irregularities occur. Figure blubb shows the modified HedgeHog activity logger that has been extended with a capacitive proximity sensor. The wristband has two electrodes, a sensing electrode underneath a slightly bigger shield electrode.

@@

The classifier was trained with various features that were extracted from windows with the length of 1 second. It turned out that greater window sizes do not provide better results. In order to suppress noise contained in the capacitive proximity sensing data, we applied a moving average filter with a kernel size of 10. The final feature set contained the arithmetic mean, min, max, median and standard variance of both accelerometer and proximity data. These simple feature types represent standard features applied in activity recognition.

@@

\begin{table}
	\caption{Comparison of two Confusion matrices for an exemplary user. The left part includes the capacitive proximity sensor, the right part represents classification results without the sensor.}

@@

	\begin{scriptsize}
	\begin{tabular}%{p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{2.2cm}}
	{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
		\multicolumn{9}{c}{with proximity} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{without proximity} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
		\textbf{10} & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 	&& \textbf{9} & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 		&& a = open door	\\ 
		2 & \textbf{259} & 4 & 2 & 4 & 2 & 4 & 0 & 0 	&& 1 & \textbf{248} & 5 & 6 & 4 & 2 & 5 & 0 & 0 	&& b = sitting		\\ 
		0 & 5 & \textbf{213} & 2 & 2 & 1 & 8 & 0 & 4 	&& 0 & 3 & \textbf{212} & 3 & 2 & 1 & 6 & 0 & 8 	&& c = lying		\\ 
		3 & 5 & 2 & \textbf{52} & 9 & 3 & 1 & 1 & 1 	&& 0 & 8 & 2 & \textbf{44} & 7 & 8 & 1 & 1 & 3 		&& d = get things	\\ 
		1 & 9 & 3 & 13 & \textbf{46} & 6 & 3 & 0 & 0 	&& 1 & 7 & 1 & 11 & \textbf{37} & 9 & 2 & 0 & 10 	&& e = make bread	\\ 
		0 & 5 & 1 & 1 & 5 & \textbf{62} & 5 & 0 & 1 	&& 0 & 3 & 2 & 3 & 8 & \textbf{58} & 5 & 0 & 1 		&& f = eating		\\ 
		0 & 8 & 4 & 2 & 1 & 5 & \textbf{68} & 0 & 0 	&& 0 & 8 & 3 & 0 & 6 & 4 & \textbf{67} & 0 & 1 		&& g = drinking		\\
		1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{12} & 0 	&& 1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{13} & 0 		&& h = walking		\\
		1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & \textbf{325} 	&& 0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & \textbf{312} 	&& i = sleeping		\\ 
		\cline{1-9} \cline{11-19} 
	\end{tabular}
	\end{scriptsize}

@@

We would like to thank our evaluation participants from TU Darmstadt and Fraunhofer IGD.

>>>>

Sensing a persons activity is being researched for several years now, raising even more interest in it by the advanced technology of mobile phones, which include multiple sensors and therefore enable activity recognition with such a device \cite{brezmes2009activity}. Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors embedded in a mobile device \cite{brezmes2009activity} or attached to the body \cite{Ravi2005} to identify the activities performed. Usually, single sensor modalities are used or duplicated to detect the activities. However, it is a great challenge to identify fine-grained activities just by using a single modality like the accelerometer.
Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show how an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living.
% For this we obtained a wrist-worn activity data logger\footnote{HedgeHog Activity Logger -- \url{http://www.ess.tu-darmstadt.de/hedgehog}} and integrate a capacitive proximity sensor unobtrusively into the wristband.
For this we obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrate a capacitive proximity sensor unobtrusively into the wristband.

@@

\begin{figure}
	\centering
 		\includegraphics[trim=1cm 1cm 1cm 1cm,clip,width=\textwidth]{Images/pseudocode.pdf}
	\caption{Measurement procedure carried out by the HedgeHog sensor: using the microcontroller's Timer0 module in counting mode, the oscillating signal generated by the capacitive sensor circuit can be measured by counting the frequency pulses over a predefined gate time of approximately 9.5ms.}
	\label{fig:pseudocode}
\end{figure}

Figure \ref{fig:pseudocode} illustrates the operations required for a measurement cycle. The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms. During that counting phase, the microcontroller is sent to sleep in order to reduce power consumption. In the following, the HedgeHog applies run-length encoding on the measured data to reduce overhead and periodically logs the data to the integrated microSD card.

@@

The experiment setup aims to depict a typical scenario in daily life. Especially in the field of Ambient Assisted Living (AAL), it is desired to monitor activities like drinking, preparing lunch and sleeping. A fine-grained monitoring of such activities may help elderly or people suffering from mental diseases to maintain a healthy day/night rhythm and take action if irregularities occur. Figure \ref{fig:cap_sensor} shows the modified HedgeHog activity logger that has been extended with a capacitive proximity sensor. The wristband has two electrodes, a sensing electrode underneath a slightly bigger shield electrode.

@@

% \begin{table}
% 	\centering
% 	\caption{Some details on the participants and the activities which were performed during the experiment.}
% 	\begin{tabular}{rp{.2cm}cc}
% 		number of participants	& & 7		&	\\
% 		average age 			& & ?		&	\\
% 		male-to-female ratio 	& & 6/1 	&	\\
% 	\end{tabular}
% 	\hspace*{2cm}
% 	\begin{tabular}{lc}
% 		performed activities	& 	\\ \toprule
% 		open door				& 	\\
% 		walking					& 	\\
% 		sitting					& 	\\
% 		lying					& 	\\
% 		get things				& 	\\
% 		making bread			& 	\\
% 		eating					& 	\\
% 		drinking 				& 	\\
% 		sleeping				& 	\\
% 	\end{tabular}
% \end{table}

The classifier was trained with basic features that were extracted from a sliding window of 1 second width. Our first tests have showed that greater window sizes do not provide better results. In order to suppress noise contained in the capacitive proximity sensing data, we applied a moving average filter with a kernel size of 10. The final feature set contained the arithmetic mean, min, max, median and standard variance for each accelerometer axis and the capacitive proximity signal. These simple feature types represent standard features applied in activity recognition.

@@

\begin{table}
	\caption{Activity recognition evaluation revealing the positive impact of the capacitive proximity sensor. Here, we are comparing SVM classification presented as confusion matrices for an exemplary user, without the proximity data on the left, and with the proximity data on the right.}

@@

	\begin{scriptsize}
	\begin{tabular}%{p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}p{0.48cm}|p{2.2cm}}
	{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
		\multicolumn{9}{c}{without proximity data} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{with proximity data} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
% 		\textbf{10} & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 	&& \textbf{9} & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 		&& a = open door	\\ 
% 		2 & \textbf{259} & 4 & 2 & 4 & 2 & 4 & 0 & 0 	&& 1 & \textbf{248} & 5 & 6 & 4 & 2 & 5 & 0 & 0 	&& b = sitting		\\ 
% 		0 & 5 & \textbf{213} & 2 & 2 & 1 & 8 & 0 & 4 	&& 0 & 3 & \textbf{212} & 3 & 2 & 1 & 6 & 0 & 8 	&& c = lying		\\ 
% 		3 & 5 & 2 & \textbf{52} & 9 & 3 & 1 & 1 & 1 	&& 0 & 8 & 2 & \textbf{44} & 7 & 8 & 1 & 1 & 3 		&& d = get things	\\ 
% 		1 & 9 & 3 & 13 & \textbf{46} & 6 & 3 & 0 & 0 	&& 1 & 7 & 1 & 11 & \textbf{37} & 9 & 2 & 0 & 10 	&& e = make bread	\\ 
% 		0 & 5 & 1 & 1 & 5 & \textbf{62} & 5 & 0 & 1 	&& 0 & 3 & 2 & 3 & 8 & \textbf{58} & 5 & 0 & 1 		&& f = eating		\\ 
% 		0 & 8 & 4 & 2 & 1 & 5 & \textbf{68} & 0 & 0 	&& 0 & 8 & 3 & 0 & 6 & 4 & \textbf{67} & 0 & 1 		&& g = drinking		\\
% 		1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{12} & 0 	&& 1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{13} & 0 		&& h = walking		\\
% 		1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & \textbf{325} 	&& 0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & \textbf{312} 	&& i = sleeping		\\ 
		\textbf{9} & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 		&& \textbf{10} & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 		&& a = open door	\\ 
		1 & \textbf{248} & 5 & 6 & 4 & 2 & 5 & 0 & 0 	&& 2 & \textbf{259} & 4 & 2 & 4 & 2 & 4 & 0 & 0 	&&  b = sitting		\\ 
		0 & 3 & \textbf{212} & 3 & 2 & 1 & 6 & 0 & 8 	&& 0 & 5 & \textbf{213} & 2 & 2 & 1 & 8 & 0 & 4 	&& c = lying		\\ 
		0 & 8 & 2 & \textbf{44} & 7 & 8 & 1 & 1 & 3 	&& 3 & 5 & 2 & \textbf{52} & 9 & 3 & 1 & 1 & 1 		&& d = get things	\\ 
		1 & 7 & 1 & 11 & \textbf{37} & 9 & 2 & 0 & 10 	&& 1 & 9 & 3 & 13 & \textbf{46} & 6 & 3 & 0 & 0 	&& e = make bread	\\ 
		0 & 3 & 2 & 3 & 8 & \textbf{58} & 5 & 0 & 1 	&& 0 & 5 & 1 & 1 & 5 & \textbf{62} & 5 & 0 & 1 		&& f = eating		\\ 
		0 & 8 & 3 & 0 & 6 & 4 & \textbf{67} & 0 & 1 	&& 0 & 8 & 4 & 2 & 1 & 5 & \textbf{68} & 0 & 0 		&& g = drinking		\\
		1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{12} & 0 	&& 1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{13} & 0 		&& h = walking		\\
		0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & \textbf{312} 	&& 1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & \textbf{325} 	&& i = sleeping		\\ 
		\cline{1-9} \cline{11-19} 
	\end{tabular}
	\end{scriptsize}

@@

Anonymized for review.

% We would like to thank our evaluation participants from TU Darmstadt and Fraunhofer IGD.
